Modal Abstractions for Virtualizing Memory Addresses

### 2024-11-13 examples.tex
compression, clarifying relation to syeda et al
more edits, adjusting presentation of some code figures, and removing some redundant text now has the experiments nearly fitting in 20 pages, but then there is related work after that still

\newcommand{\fpaddr}{\texttt{fpaddr}}
\newcommand{\specline}[1]{{\color{blue}\left\{#1\right\}}}
\newcommand{\sumapacesfull}[2]{
  \ownGhost\gammaPreds{\authfull{\singletonMap{#1}{#2}}}
}
\section{Experiments}
\label{sec:experiment}
To both validate and demonstrate the value of the modal approach to reasoning about virtual memory management, we study several
distillations of real concerns of virtual memory managers.
Recall from Section \ref{sec:logic} that virtual points-to assertions work just like regular points-to assertions, by design.
In this section we work through two critical and challenging aspects of virtual memory management.
First, in several stages, we work up to mapping a new page in the current address space.
This requires a number of challenging substeps: dynamically traversing a page table to find
the appropriate L1 entry to update; inserting additional levels of the page table if necessary (updating
the VMM invariants along the way);
converting the physical addresses found in intermediate entries into the corresponding virtual addresses
that can be used for access;
installing the new mapping;
and collecting sufficient resources to form a virtual points-to assertion.
Of these, only the second-to-last step (installing the correct mapping into the
current address space) has previously been formally verified with respect to a machine model with address translation.
Second, we formally verify a switch into a new address space as part of a task switch.\todo{Colin will double-check the TLB work's details before claiming this is the first}

@@

Note that none of the verification for this function assumes specific page table levels --- logical parameter \textsf{v} represents the level
of the entry passed as an argument, and this code
is used for all three level transitions when traversing page tables (4 to 3, 3 to 2, 2 to 1).
This comes into play with a subtlety of the specification of \lstinline|pte_get_next_table| that we will
revisit several times: \lstinline|pte_get_next_table|'s specification
assumes it is given a virtual \emph{vpte-pointsto}\footnote{A \emph{vpte-pointsto}
$\vaddr\mapsto_{\textsf{vpte,qfrac}} \textsf{\paddr~\val}$ is a virtual points-to granting
access to virtual address $\vaddr$, which is known to hold \textsf{\val}, but additionally exposing the physical address
$\paddr$ that the virtual address translates to. This is commonly used, as the name suggests, when updating PTEs,
where we require assurance that we are accessing the intended PTE.
Its definition is just like the regular virtual points-to in Figure \ref{fig:virtualpointstosharing},
except taking the physical address as a parameter rather than existentially quantifying it.
}
 granting access to the specified entry,
but its postcondition does not yield new virtual points-to assertions!
Instead it merely computes the base address of the next table, and returns adequate capabilities (discussed in Section \ref{subsec:identitymappings})
for the \emph{caller} to construct a vpte-pointsto for the next table level (if this is not an L1 entry ---
the caller knows which level of the table this is for).
\looseness=-1

@@

@@

@@

If the entry is not set, \textsf{pte\_initialize} is called (Line \ref{line:call_to_pte_initialize} in Figure \ref{fig:calltopteinitialize}) 
for a physical page (utilizing the page-allocator's \textsf{kalloc} -- currently the only axiomatized call in the the proof of \textsf{pte\_initialize} 
(Line \ref{line:call_to_kalloc} in Figure \ref{pteinitializespec}). 
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{r}} \textsf{\_} \ast \lnot(\textsf{entry\_present entry\_val} }_{\rtv}$
$\specline{\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
;;void pte_initialize(pte_t *entry) {
... ;;set up the stack    
;;allocate a full zeroed page for 512 8-byte entries 
callq  81 <kalloc>	;;pte_t *local = kalloc();(*\label{line:call_to_kalloc}*)
mov    rax,-0x10[rbp] ;; Store into 'local'
;;entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);
mov    -0x10[rbp],rax
mov    -0x8[rbp],rdi
mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE} }_{\rtv}$
$\specline{\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
$\specline{\texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast \lnot(\textsf{entry\_present entry\_val}) }$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entry} \; \textsf{pfn\_set(entryv nextpaddr)}}$
$\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$
... ;;clean up the stack, return
\end{lstlisting}
\vspace{-1em}
\caption{Allocating a physical page }
\label{pteinitializespec}
\end{figure}

@@

@@

\subsection{Change of Address Space}
A critical piece of \emph{trusted} code in current verified OS kernels is the assembly code to change the current address space; current verified OS kernels currently 
lack effective ways to specify and reason about this low-level operation, for reasons outlined in Section \ref{sec:relwork}.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
;; Assume the save-space is in rdi, load-space in rsi. First, save the yielding context
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\Xi,m) \ast [\rtv'](\mathcal{I}\texttt{ASpace}(\theta',\Xi',m') \ast \texttt{Pother}) \ast \texttt{rsi}\mapsto_{\textsf{r}} new \ast \texttt{rdi}\mapsto_{\textsf{r}} old \ast \texttt{rbx}\mapsto_{\textsf{r}} \texttt{rbxv} }_{\rtv}$
$\specline{\texttt{rsp}\mapsto_{\textsf{r}} \texttt{rspv} \ast \texttt{rbp}\mapsto_{\textsf{r}} \texttt{rbpv} \ast \texttt{r12}\mapsto_{\textsf{r}} \texttt{r12v} \ast \texttt{r13}\mapsto_{\textsf{r}} \texttt{r13v} \ast \texttt{r14}\mapsto_{\textsf{r}} \texttt{r14v} \ast \texttt{r15}\mapsto_{\textsf{r}} \texttt{r15v}}_{\rtv}$
$\specline{\mathsf{ContextAt}(old,\_) \ast \mathsf{ContextAt}(new,[\mathsf{rbxv}',\ldots,\rtv']) }_{\rtv}$
mov 0[rdi], rbx
... ;; mov rsp, rbp, r12, r13, r14, saved to offsets 8, 16, 24, 32, and 40 from rdi
mov 48[rdi], r15
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\Xi,m) \ast [\rtv'](\mathcal{I}\texttt{ASpace}(\theta',\Xi',m') \ast \texttt{Pother})  \ast  \texttt{rsi}\mapsto_{\textsf{r}} new \ast \texttt{rdi}\mapsto_{\textsf{r}} old \ast \texttt{rbx}\mapsto_{\textsf{r}} \texttt{rbxv} }_{\rtv}$
$\specline{ \texttt{rsp}\mapsto_{\textsf{r}} \texttt{rspv} \ast \texttt{rbp}\mapsto_{\textsf{r}} \texttt{rbpv} \ast \texttt{r12}\mapsto_{\textsf{r}} \texttt{r12v} \ast \texttt{r13}\mapsto_{\textsf{r}} \texttt{r13v} \ast \texttt{r14}\mapsto_{\textsf{r}} \texttt{r14v} \ast \texttt{r15}\mapsto_{\textsf{r}} \texttt{r15v}}_{\rtv}$
$\specline{\mathsf{ContextAt}(old,[\mathsf{rbxv,\ldots,\rtv}]) \ast \mathsf{ContextAt}(new,[\mathsf{rbxv}',\ldots,\rtv']) }_{\rtv}$
mov 56[%rdi], %cr3
$\specline{\ldots \ast \texttt{rdi+56} \mapsto_{\textsf{v}} \rtv}_{\rtv}$    
;; Restore target context
mov rbx, 0[rsi] 
mov rsp, 8[rsi] ;; Switch to new stack, which may not be mapped in the current address space!
... ;; load rbp, r12, r13, r14, from offsets 16, 24, 32, and 40 from rsi
mov r15, 48[rsi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\Xi,m) \ast [\rtv'](\mathcal{I}\texttt{ASpace}(\theta',\Xi',m') \ast \texttt{Pother})  \ast  \texttt{rsi}\mapsto_{\textsf{r}} new \ast \texttt{rdi}\mapsto_{\textsf{r}} old \ast \texttt{rbx}\mapsto_{\textsf{r}} \texttt{rbxv}' }_{\rtv}$
$\specline{\texttt{rsp}\mapsto_{\textsf{r}} \texttt{rspv}' \ast \texttt{rbp}\mapsto_{\textsf{r}} \texttt{rbpv}' \ast \texttt{r12}\mapsto_{\textsf{r}} \texttt{r12v}' \ast \texttt{r13}\mapsto_{\textsf{r}} \texttt{r13v}' \ast \texttt{r14}\mapsto_{\textsf{r}} \texttt{r14v}' \ast \texttt{r15}\mapsto_{\textsf{r}} \texttt{r15v}'}_{\rtv}$
$\specline{\mathsf{ContextAt}(old,[\mathsf{rbxv,\ldots,\rtv}]) \ast \mathsf{ContextAt}(new,[\mathsf{rbxv}',\ldots,\rtv']) }_{\rtv}$
;; Switch to the new address space
mov cr3, 56[rsi]
$\specline{ [\rtv](\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\Xi,m) \ast \mathsf{ContextAt}(old,[\mathsf{rbxv,\ldots,\rtv}]) \ast \mathsf{ContextAt}(new,[\mathsf{rbxv}',\ldots,\rtv']))  }_{\rtv}$
$\specline{ \mathcal{I}\texttt{ASpace}(\theta',\Xi',m') \ast \texttt{Pother} \ast  \texttt{rsi}\mapsto_{\textsf{r}} new \ast \texttt{rdi}\mapsto_{\textsf{r}} old \ast \ldots }_{\rtv'}$
\end{lstlisting}
\vspace{-1em}
\caption{Basic task switch code that switches address spaces.}
\label{fig:swtch}
\end{figure}

>>>>

\newcommand{\fpaddr}{\texttt{fpaddr}}
\newcommand{\specline}[1]{{\color{blue}\left\{#1\right\}}}
\newcommand{\sumapacesfull}[2]{
  \ownGhost\gammaPreds{\authfull{\singletonMap{#1}{#2}}}
}
\section{Experiments}
\label{sec:experiment}
%To both validate and demonstrate the value of the modal approach to reasoning about virtual memory management, 
% we study several
% We validate our logic by studying
% distillations of key VMM functionality.
% real concerns of virtual memory managers.
% Recall from Section \ref{sec:logic} that virtual points-to assertions work just like regular points-to assertions, by design.
In this section we verify several critical and challenging pieces of VMM code.
First, in several stages, we work up to mapping a new page in the current address space.
This requires a number of independently challenging substeps: dynamically traversing a page table to find
the appropriate L1 entry to update; inserting additional levels of the page table if necessary (updating
the VMM invariants along the way);
converting the physical addresses found in intermediate entries into the corresponding virtual addresses
that can be used for access;
installing the new mapping;
and collecting sufficient resources to form a virtual points-to assertion.
Of these, only the second-to-last step (installing the correct mapping into the
current address space) has previously been formally verified with respect to a machine model with address translation.
Second, we formally verify a switch into a new address space as part of a task switch,
the first such verification handling both old and new processes' assertions at the time of the switch.

@@

Note that none of the verification for this function assumes specific page table levels --- logical parameter \textsf{v} represents the level
of the entry passed as an argument, and this code
is used for all three level transitions when traversing page tables (4 to 3, 3 to 2, 2 to 1).
This comes into play with a subtlety of the specification of \lstinline|pte_get_next_table| that we will
revisit several times: \lstinline|pte_get_next_table|'s specification
assumes it is given a virtual \emph{vpte-pointsto}
% \footnote{A \emph{vpte-pointsto}
% $\vaddr\mapsto_{\textsf{vpte,qfrac}} \textsf{\paddr~\val}$ is a virtual points-to granting
% access to virtual address $\vaddr$, which is known to hold \textsf{\val}, but additionally exposing the physical address
% $\paddr$ that the virtual address translates to. This is commonly used, as the name suggests, when updating PTEs,
% where we require assurance that we are accessing the intended PTE.
% Its definition is just like the regular virtual points-to in Figure \ref{fig:virtualpointstosharing},
% except taking the physical address as a parameter rather than existentially quantifying it.
% }
(a virtual points-to exposing the underlying physical address; see Section \ref{sec:mapnew}) granting access to the specified entry,
but its postcondition does not yield new virtual points-to assertions!
Instead it merely computes the base address of the next table, and returns adequate capabilities (discussed in Section \ref{subsec:identitymappings})
for the \emph{caller} to construct a vpte-pointsto for the next table level (if this is not an L1 entry ---
the caller knows which level of the table this is for).
\looseness=-1

@@

\ifPLDI
;;pte_t *pte_get_next_table(pte_t *entry) { ... // setting up the stack for a pte_t *next elided
mov    -0x8[rbp],rdi $\specline{\ldots \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ldots }$
mov     rdi, r8 $\specline{\ldots \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ldots }$
mov    [r8],rdi (*\label{line:read_entry_contents}*) $\specline{\ldots \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  \ast\ldots}$
\else

@@

\fi

@@

If the entry is not set, \textsf{pte\_initialize} is called (Line \ref{line:call_to_pte_initialize} in Figure \ref{fig:calltopteinitialize}) 
for a physical page (utilizing the page-allocator's \textsf{kalloc} -- currently the only axiomatized call in the the proof of \textsf{pte\_initialize} 
(Line \ref{line:call_to_kalloc} in Figure \ref{pteinitializespec}). 
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{r}} \textsf{\_} \ast \lnot(\textsf{entry\_present entry\_val} }_{\rtv}$
$\specline{\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,1}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
;;void pte_initialize(pte_t *entry) {
... ;;set up the stack    
;;allocate a full zeroed page for 512 8-byte entries 
callq  81 <kalloc>	;;pte_t *local = kalloc();(*\label{line:call_to_kalloc}*)
mov    rax,-0x10[rbp] ;; Store into 'local'
;;entry->pfn = PTE_ADDR_TO_PFN((uintptr_t) local);
mov    -0x10[rbp],rax; mov    -0x8[rbp],rdi; mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry+KERNBASE} }_{\rtv}$
$\specline{\textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,}} \textsf{entry entryv} \ast \ghostmaptoken{\textsf{id}}{(\mathsf{entry})}{\textsf{level}} }$
$\specline{\texttt{rax} \mapsto_{\textsf{r}} \textsf{nextpaddr} \ast \lnot(\textsf{entry\_present entry\_val}) }$
$\specline{ \textsf{entry+KERNBASE} \mapsto_{\mathsf{vpte,1}} \mathsf{entry} \; \textsf{pfn\_set(entryv nextpaddr)}}$
$\specline{\ghostmaptoken{\delta{}s}{\rtv}{\delta} \ast \begin{array}{l} \ulcorner\textsf{entry\_present (pte\_initialize(pfn\_set(entry\_val nextpaddr)),level)}\urcorner \wand \\  \;\;\;\;\;\;\; \forall_{i\in\textsf{0..511}} \; \textsf{table\_root (pte\_initialized (pfn\_set (entry\_val nextpaddr))) + i * 8} \mapsto_{\textsf{id}} \textsf{level-1}    \end{array}  }_{\rtv}$
... ;;clean up the stack, return
\end{lstlisting}
\vspace{-1em}
\caption{Allocating a physical page }
\label{pteinitializespec}
\end{figure}

@@

\label{sec:mapnew}

@@

\subsection{Change of Address Space}
A critical piece of \emph{trusted} code in current verified OS kernels is the assembly code to change the current address space; current verified OS kernels currently 
lack effective ways to specify and reason about this low-level operation, for reasons outlined in Section \ref{sec:relwork}.
\begin{figure}\footnotesize
\begin{lstlisting}[mathescape,escapeinside={(*}{*)}]
;; Assume the save-space is in rdi, load-space in rsi. First, save the yielding context
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\Xi,m) \ast [\rtv'](\mathcal{I}\texttt{ASpace}(\theta',\Xi',m') \ast \texttt{Pother}) \ast \texttt{rsi}\mapsto_{\textsf{r}} new \ast \texttt{rdi}\mapsto_{\textsf{r}} old \ast \texttt{rbx}\mapsto_{\textsf{r}} \texttt{rbxv} }_{\rtv}$
$\specline{\texttt{rsp}\mapsto_{\textsf{r}} \texttt{rspv} \ast \texttt{rbp}\mapsto_{\textsf{r}} \texttt{rbpv} \ast \texttt{r12}\mapsto_{\textsf{r}} \texttt{r12v} \ast \texttt{r13}\mapsto_{\textsf{r}} \texttt{r13v} \ast \texttt{r14}\mapsto_{\textsf{r}} \texttt{r14v} \ast \texttt{r15}\mapsto_{\textsf{r}} \texttt{r15v}}_{\rtv}$
$\specline{\mathsf{ContextAt}(old,\_) \ast \mathsf{ContextAt}(new,[\mathsf{rbxv}',\ldots,\rtv']) }_{\rtv}$
mov 0[rdi], rbx ... ;; also mov rsp, rbp, r12, r13, r14, saved to offsets 8, 16, 24, 32, and 40 from rdi
mov 48[rdi], r15
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\Xi,m) \ast [\rtv'](\mathcal{I}\texttt{ASpace}(\theta',\Xi',m') \ast \texttt{Pother})  \ast  \texttt{rsi}\mapsto_{\textsf{r}} new \ast \texttt{rdi}\mapsto_{\textsf{r}} old \ast \texttt{rbx}\mapsto_{\textsf{r}} \texttt{rbxv} }_{\rtv}$
$\specline{ \texttt{rsp}\mapsto_{\textsf{r}} \texttt{rspv} \ast \texttt{rbp}\mapsto_{\textsf{r}} \texttt{rbpv} \ast \texttt{r12}\mapsto_{\textsf{r}} \texttt{r12v} \ast \texttt{r13}\mapsto_{\textsf{r}} \texttt{r13v} \ast \texttt{r14}\mapsto_{\textsf{r}} \texttt{r14v} \ast \texttt{r15}\mapsto_{\textsf{r}} \texttt{r15v}}_{\rtv}$
$\specline{\mathsf{ContextAt}(old,[\mathsf{rbxv,\ldots,\rtv}]) \ast \mathsf{ContextAt}(new,[\mathsf{rbxv}',\ldots,\rtv']) }_{\rtv}$
mov 56[%rdi], %cr3 $\specline{\ldots \ast \texttt{rdi+56} \mapsto_{\textsf{v}} \rtv \ast \ldots }_{\rtv}$    
;; Restore target context
mov rbx, 0[rsi] 
mov rsp, 8[rsi] ;; Switch to new stack, which may not be mapped in the current address space!
... ;; load rbp, r12, r13, r14, from offsets 16, 24, 32, and 40 from rsi
mov r15, 48[rsi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\Xi,m) \ast [\rtv'](\mathcal{I}\texttt{ASpace}(\theta',\Xi',m') \ast \texttt{Pother})  \ast  \texttt{rsi}\mapsto_{\textsf{r}} new \ast \texttt{rdi}\mapsto_{\textsf{r}} old \ast \texttt{rbx}\mapsto_{\textsf{r}} \texttt{rbxv}' }_{\rtv}$
$\specline{\texttt{rsp}\mapsto_{\textsf{r}} \texttt{rspv}' \ast \texttt{rbp}\mapsto_{\textsf{r}} \texttt{rbpv}' \ast \texttt{r12}\mapsto_{\textsf{r}} \texttt{r12v}' \ast \texttt{r13}\mapsto_{\textsf{r}} \texttt{r13v}' \ast \texttt{r14}\mapsto_{\textsf{r}} \texttt{r14v}' \ast \texttt{r15}\mapsto_{\textsf{r}} \texttt{r15v}'}_{\rtv}$
$\specline{\mathsf{ContextAt}(old,[\mathsf{rbxv,\ldots,\rtv}]) \ast \mathsf{ContextAt}(new,[\mathsf{rbxv}',\ldots,\rtv']) }_{\rtv}$
mov cr3, 56[rsi] ;; <-- Switch to the new address space
$\specline{ [\rtv](\textsf{P} \ast \mathcal{I}\texttt{ASpace}(\theta,\Xi,m) \ast \mathsf{ContextAt}(old,[\mathsf{rbxv,\ldots,\rtv}]) \ast \mathsf{ContextAt}(new,[\mathsf{rbxv}',\ldots,\rtv']))  }_{\rtv}$
$\specline{ \mathcal{I}\texttt{ASpace}(\theta',\Xi',m') \ast \texttt{Pother} \ast  \texttt{rsi}\mapsto_{\textsf{r}} new \ast \texttt{rdi}\mapsto_{\textsf{r}} old \ast \ldots }_{\rtv'}$
\end{lstlisting}
\vspace{-1em}
\caption{Basic task switch code that switches address spaces.}
\label{fig:swtch}
\end{figure}
